---
title: "Introductory econometrics - Illustrations"
subtitle: "Free to use, happy to receive a mention if you find this useful for your class."
date: "Winter 22/23"
author: Simon Heß
fontsize: 10pt
classoption: compress
output:
  bookdown::beamer_presentation2:
    slide_level: 3
    theme: metropolis
    highlight: haddock
    keep_tex: yes
    toc: no
---


```{r, echo = FALSE} 
# code that I use to make beamer-overlay slides with r
hook_plot <- knitr::knit_hooks$get("plot")
knitr::knit_hooks$set(
  plot = function(x, options) {
    if (is.null(options$overlay.plot)) {
      return(hook_plot(x, options))
    } else {
      i <- options$fig.cur
      offset <- 0
      if (is.null(options$offset)) {
        overlay_offset <- 0
      } else {
        overlay_offset <- options$offset
      } 
      hand <- which(i==options$overlay.plot)
      bf <- paste0("\\only<", i+overlay_offset, "| handout:", ifelse(is.null(hand),0,hand), ">{")
      paste(c(bf, knitr::hook_plot_tex(x, options), "}"), collapse = "")
    }
  }
)
```



# Sampling distribution of OLS

```{r fig-sd, fig.width = 9, fig.height = 6, out.width = "\\textwidth", echo = FALSE, overlay.plot = 1}
library(scales)
set.seed(2)
par(mar = c(2, 0.1, 3, 0))
pop <- 5000             #population
x <- runif(pop) * 2
y <- x + rnorm(pop) / 2
samples <- c(rep(0.006, 4000))
olslines <- list()
par(mfrow = c(1, 2))
show_examples <- c(1:10, 25, 50, 100, 500, 1000, 5000, 10000)
beta1 <- c()

for (i in seq_along(samples)) {
  sample_i <- sample(1:pop, pop * samples[i])
  ols <- lm(y[sample_i] ~ x[sample_i])
  olslines <- c(list(ols), olslines)
  beta1 <- c(beta1, coefficients(ols)[2])
  if (i %in% show_examples | i == length(samples)) {
    plot(x, y,
      col = alpha("grey", 0.1),
      pch = 19,
      xlim = c(.05, 2.05),
      ylim = c(.4, 2.4),
      main = paste0(
        signif(length(olslines), 3), " line fits on ",
        signif(length(olslines), 3)," samples"
      ),
      xaxt = 'n',
      yaxt = 'n'
    )
    
    for (olsold in olslines) {
      abline(
        reg = olsold,
        col = alpha("darkslategray", 1 / sqrt(sqrt(i))),
        lwd = 1.5
      )
    }
    if (i < length(samples)) {
      points(
        x[sample_i],
        y[sample_i],
        col = alpha("indianred", 1),
        pch = 19,
        xlim = 1 + c(-1.05, 1.05),
        ylim = 1 + c(-1.4, 1.4)
      )
      abline(reg = ols,
             col = alpha("indianred", 1),
             lwd = 5)
    }
    abline(
      a = 0,
      b = 1,
      lwd = 3,
      col = "black"
    )
    hist(
      beta1,
      breaks = 0.3 + 1.4 * (0:30) / 30,
      main = paste0("histrogram of beta 1 across ", signif(length(olslines), 3), " samples"),
      xlim = c(0.4, 1.6),
      freq = F,
      yaxt = 'n',
      col = "darkslategray"
    )
  }
}

```


# Sampling distribution of OLS under heteroskedasticity


```{r fig-heteroscedacisity1, fig.width = 9, fig.height = 6, out.width = "\\textwidth", echo = FALSE, overlay.plot = 1}
library(scales)

set.seed(1234)
par(mfrow = c(2, 2), mar = c(1.5, 2, 2, 1))

n <- 1000
x <- runif(n) ^ 1.7

error <- rnorm(n)

error_homo <- (runif(n) - 0.5) * ((abs(x)) ^ 0) + error / 5
error_hetero <- (runif(n) - 0.5) * ((abs((x)) ^ 4)) + error / 100

# make sure they have the same unconditional variance
error_homo <- error_homo / sd(error_homo) / 2.5
error_hetero <- error_hetero / sd(error_hetero) / 2.5

y_homo <- 0.5 * x + error_homo
y_hetero <- 0.5 * x + error_hetero

samples <- rep(0.03, 1000) #the different samples to draw

olslines_homo <- list() #where to save the line fits
olslines_hetero <- list() #where to save the line fits
show_examples <- c(1:10, 25, 50, 100, 500, 1000, 5000, 10000)

for (i in seq_along(samples)) {
  sample_i <- sample(1:n, n * samples[i])
  ols_homo <- lm(y_homo[sample_i] ~ x[sample_i])
  ols_hetero <- lm(y_hetero[sample_i] ~ x[sample_i])
  cols <- ifelse(1:n %in% sample_i, "indianred", "grey50")
  alphas <- ifelse(1:n %in% sample_i, 1, 0.4)
  if (i %in% c(show_examples, length(samples))) {
    
    ##### plot 1 ######
    plot(
      x, y_homo,
      col = alpha(cols, alphas), pch = 19,
      xlim = c(-0.1, 1.1), ylim = c(-0.5, 1.5),
      main = paste0(
        "OLS on sample from pop with var(e)=",
        round(var(error_homo), 4),
        "=var(e|x)"
      ),
      ylab = "one sample"
    )
    abline(a = 0, b = 0.5, lwd = 3, col = "grey50")
    abline(ols_homo, col = "indianred", lwd = 3)
    
    ##### plot 2 ######
    plot(
      x,
      y_hetero,
      col = alpha(cols, alphas), pch = 19,
      xlim = c(-0.1, 1.1), ylim = c(-0.5, 1.5),
      main = paste0(
        "OLS  on sample from pop with var(e)=",
        round(var(error_hetero), 4),
        " != var(e|x)"
      ),
      ylab = paste0(i, " samples")
    )
    abline(a = 0, b = 0.5, lwd = 3, col = "grey50")
    abline(ols_hetero, col = "indianred", lwd = 3)
  }
  
  olslines_homo <- c(list(ols_homo), olslines_homo)
  olslines_hetero <- c(list(ols_hetero), olslines_hetero)
  
  if (i %in% c(show_examples, length(samples))) {
    # plot 3
    plot(
      x, y_homo,
      col = alpha("grey", 0.1), pch = 19,
      xlim = c(-0.1, 1.1), ylim = c(-0.4, 1.4)
    )
    for (ols in olslines_homo)
      abline(reg = ols, col = alpha("darkslategray", 3 / log(1 + i)),lwd = 1)
    abline(a = 0, b = 0.5, lwd = 3, col = "grey50")
    
    #plot 4
    plot(
      x, y_hetero,
      col = alpha("grey", 0.1), pch = 19,
      xlim = c(-0.1, 1.1), ylim = c(-0.4, 1.4)
    )
    for (ols in olslines_hetero)
      abline(reg = ols, col = alpha("darkslategray", 3 / log(1 + i)), lwd = 1)
    abline(a = 0, b = 0.5, lwd = 3, col = "grey50")
  }
}

```

# Measurement error
## Measurement error in $y$ 

```{r fig-yerror, fig.width = 10, fig.height = 7, out.width = "\\textwidth", echo = FALSE, overlay.plot = 1}
library(scales)
set.seed(123)
n<-200
x <- runif(n)
error<-rnorm(n)
y <- x 
errorfactor <- c(0:10)/20
for (factor in errorfactor) {
  y0<-y+factor*error
  ols<-lm(y0~x)
  plot(x, y0, col = alpha("indianred",0.8), pch = 19, xlim = c(-0.5,1.5), ylim = c(-0.5,2.5))
  text(x=0,y=2.5,adj=c(0,1),paste0(  "b_0=",round(ols$coefficients[1],2), "   b_1=",   round(ols$coefficients[2] ,2) ))
  abline(0,1, col="gray", lwd = 3)
  abline(ols, col="indianred", lwd = 3)
}
```
## Measurement error in $x$ 

```{r fig-xerror, fig.width = 10, fig.height = 7, out.width = "\\textwidth", echo = FALSE, overlay.plot = 1}
set.seed(123)

n<-200
x <- runif(n)
error<-rnorm(n)
y <- x
errorfactor <- c(0:10)/10
for (factor in errorfactor) {
  x0<-x+factor*error
  ols<-lm(y~x0)
  plot(x0, y, col = alpha("indianred",0.8), pch = 19, xlim = c(-0.5,2), ylim = c(-0.5,2))
  text(x=0,y=2,adj=c(0,1),paste0(  "b_0=",round(ols$coefficients[1],2), "   b_1=",   round(ols$coefficients[2] ,2) ))
  abline(0,1, col="gray", lwd = 3)
  abline(lm(y~x0), col="indianred", lwd = 3)
}
```


# Truncation
## Truncated $y$ 

```{r fig-truncation, fig.width = 10, fig.height = 7, out.width = "\\textwidth", echo = FALSE, overlay.plot = 1}
set.seed(12345)
library(scales)
n<-500
x <- runif(n)
y <- x+rnorm(n)*0.2 - 0.75
truncations <- c(-Inf, -1, 0,0.125, 0.25, 0.375, 0.5, 0.627,  0.75, 0.875, 1)-0.75
for (truncation in truncations) {
  sample_i <- which(y>truncation)
  cols <- ifelse(1:n %in% sample_i, "indianred", "grey80")
  alphas <- ifelse(1:n %in% sample_i, 1, 0.4)
  plot(x, y, col = alpha(cols, alphas), pch = 19)
  abline(a=-0.75,b=1, lwd = 3, col = "grey50")
  abline(h=truncation, lwd = 3, col = "grey80")
  abline(lm(y[sample_i] ~ x[sample_i]), col="indianred", lwd = 3)
}
```

```{r eval=F}
code
```

## Truncated $x$ 

```{r fig-truncationx, fig.width = 10, fig.height = 7, out.width = "\\textwidth", echo = FALSE, overlay.plot = 1}
set.seed(12345)
library(scales)
n <- 500
x <- runif(n)
y <- x + rnorm(n) * 0.2 - 0.75
truncations <- c(-Inf, 0.125, 0.25, 0.375, 0.5, 0.627,  0.75)
for (truncation in truncations) {
  sample_i <- which(x > truncation)
  cols <- ifelse(1:n %in% sample_i, "indianred", "grey80")
  alphas <- ifelse(1:n %in% sample_i, 1, 0.4)
  plot(x, y, col = alpha(cols, alphas), pch = 19)
  abline(a = -0.75, b = 1, lwd = 3, col = "grey50")
  abline(v = truncation, lwd = 3, col = "grey80")
  abline(lm(y[sample_i] ~ x[sample_i]), col = "indianred", lwd = 3)
}
```



# Convergence Concepts
### WLLN -- Example

After each roll,  each student computes the mean of their past rolls.

This mean, $\bar{y}$, is our $y_n$, and the WLLN tells us that $y_n\overset{p}{\to}\mu=3.5$. 

```{r fig-plim, fig.width = 8, fig.height = 4, out.width = "0.95\\textwidth", echo = FALSE, overlay.plot = c(1,7)}
set.seed(12345)
library(scales)
n <- 5000
k <- 100
rolls <- matrix(round(runif(n * k) * 6 + 0.5), n, k)
rolls[1, ] <-  rolls[2, ]  ##this is actually wrong, but makes the presentation easier
means <- rolls
for (i in 2:n) 
  means[i, ] <- colMeans(rolls[1:i, ])
layout(matrix(c(1, 1, 2), nrow = 1, ncol = 3, byrow = TRUE ))
steps_to_plot <- c(2:10, 25, 50, 100, 1000, 5000)
for (showsteps in steps_to_plot) {
  par(mar = c(4, 2, 2, 0))
  matplot(
    y = means[1:showsteps, ],
    x = 1:showsteps - 1,
    type = 'l', lty = "solid", col = alpha("black", 0.1),
    main = paste("Mean of", signif(showsteps - 1, -1), "roll(s) of a die"),
    xlab = "n", xaxt = 'n',
    xlim = c(0, max(10, showsteps + 1)),
    ylab = "",
    ylim = c(0.5, 6.5)
  )
  if (showsteps > 2)
    axis(1)
  points(
    y = means[showsteps, ] + rnorm(k) * 0.04 * 2 / showsteps,
    x = -1 + rep(showsteps, k) + runif(k) * 0.2,
    col = alpha("indianred", 0.8), pch = 16, cex = 0.5
  )
  
  yhist <-
    hist(means[showsteps, ],
         plot = FALSE,
         breaks = seq(from = 0.5, to = 6.5, length.out = min(round(6 + sqrt(showsteps)), 30) )) # note: this uses probability=TRUE
  par(mar = c(4, 1, 2, 2))
  barplot(
    yhist$counts,
    axes = FALSE,
    xlim = c(0, 100),
    space = 0,
    horiz = TRUE,
    col = alpha("indianred", 0.6),
    main = "Frequency"
  ) # barplot
  axis(1)
}

```

### CLT  -- Example

After each roll,  each student computes the mean of their past rolls.

Before, we looked at $\bar{x}_n$, now we look at \textcolor{red}{$\sqrt{n}(\bar{x}_n-\mu)$}.

The CLT tells us that this RV converges to \textcolor{blue}{$N(0,\sigma)$}.


```{r fig-clt, fig.width = 8.5, fig.height = 4, out.width = "0.95\\textwidth", echo = FALSE, overlay.plot = c(1,5), echo=FALSE,warning=FALSE,message=FALSE}

# somehow this is still a bit of, because base-r does not seem to allow me to align the histogram and the line plot axis. so i have to use a hacky way which sometimes shifts by a bit.
set.seed(100)
library(scales)
library(Rfast)
n <- 30
k <- 10000
plotbandwidth <- 8.4
rolls <- matrix(round(runif(n * k) * 6 + 0.5), n, k)
means <- rolls
for (i in 2:n) 
  means[i, ] <- Rfast::colmeans(rolls[1:i, ])

means <- (means - 3.5) * sqrt(1:n)

layout(matrix(c(1, 1, 2), nrow = 1, ncol = 3, byrow = TRUE))

for (showsteps in c(1:5, 10, 15, 20, 25, 30)) {
  par(mar = c(4, 2, 2, 0))
  steps = sort(unique(means[showsteps, ]))
  stepsize <- min(steps[2:length(steps)] - steps[1:(length(steps) - 1)])
  
  matplot(
    y = matrix(means[1:showsteps, sample.int(k, round((showsteps / n) * k))], nrow =
                 showsteps),
    x = 1:showsteps - 1,
    type = 'l',
    ylim = c(-plotbandwidth, plotbandwidth),
    lty = "solid",
    col = alpha("black", 0.02),
    xlim = c(0, max(10, showsteps)),
    main = paste("sqrt(n)-scaled mean of", showsteps, "roll(s) of a die"),
    xlab = "n",
    ylab = "",
    xaxt = 'n'
  )
  if (showsteps > 2)
    axis(1)
  
  yhist <-
    hist(
      means[showsteps, ],
      plot = FALSE,
      breaks = seq(from = -plotbandwidth, to = plotbandwidth, stepsize)
    )
  yy <- seq(-plotbandwidth, plotbandwidth, 0.2)
  points(
    y = means[showsteps, ] + rnorm(k) * 0.05,
    x = -1.5 + rep(showsteps, k) + runif(k) * 0.3 + 0.4,
    cex = 0.2,
    col = alpha("indianred", 0.8),
    pch = 16
  )
  par(mar = c(4, 1, 2, 2))
  barplot(
    c(0, yhist$density),
    ylim = c(-plotbandwidth, plotbandwidth),
    width = c(-plotbandwidth + stepsize / 2, rep(stepsize, length(means[showsteps, ]))),
    xlim = c(0, max(yhist$density, dnorm(yy, sd = sqrt(
      35 / 12
    )) * 2)),
    axes = FALSE, space = 0, horiz = TRUE,
    col = alpha("indianred", 0.6),
    main = "Density"
  ) # barplot
  
  lines(dnorm(yy, sd = sqrt(35 / 12)), yy, col = "blue", lw = 4) # line
}

```


# CIs
```{r fig-ci, fig.width = 10, fig.height = 6, out.width = "\\textwidth", echo = FALSE, overlay.plot = 1}
library(scales)
set.seed(123)
par(mar = c(4, 4, 2, 1))
pop <- 5000 #population
x <- runif(pop) * 2
true_beta_1 <- 1.2
y <- true_beta_1 * x + rnorm(pop) / 2
samples <- c(rep(0.006, 100))
olslines <- list()
par(mfrow = c(1, 2))
show_examples <- sort(unique(c(1:10, 20 * (1:5))))
cis <- matrix(NA, nrow = 0, ncol = 2)

for (i in c(seq_along(samples), rep(length(samples), 10))) {
  if (nrow(cis) < length(samples)) {
    sample_i <- sample(1:pop, pop * samples[i])
    ols <- lm(yy ~ xx, data = data.frame(yy = y[sample_i], xx = x[sample_i]))
    olslines <- c(list(ols), olslines)
    cis <- rbind(cis, confint(ols)[2, ])
    
    predx <- seq(0, 2, 0.01)
    preds <- predict(ols, newdata = data.frame(xx = predx),
                     interval = 'confidence')
  }
  if (i %in% show_examples) {
    plot(x, y,
      col = alpha("grey", 0.1), pch = 19,
      main = paste0("OLS on sample. Population beta1=", true_beta_1),
      xlim = c(-0.0, 2.0), ylim =  c(-0.4, 2.4),
      xlab = "x", ylab = "y"
    )
    
    if (i != length(samples)) {
      points(x[sample_i], y[sample_i], col = alpha("indianred", 1), pch = 19)
      
      abline(reg = ols,
             col = alpha("indianred", 1),
             lwd = 5)

      # shaded CI
      polygon(c(rev(predx), predx), c(rev(preds[, 3]), preds[, 2]), col = alpha('indianred', 0.4), border = NA )
      lines(predx, preds[, 3], lty = 'dashed', col = 'black')
      lines(predx, preds[, 2], lty = 'dashed', col = 'black')
      
      text(x = 0.8, y = 0, adj = 0,
        paste0("95%-CI for beta0: [", round(confint(ols)[1, 1], 2), ",", round(confint(ols)[1, 2], 2), "]"),
        col = "indianred", cex = 1
      )
      text(x = 0.8, y = -0.25, adj = 0,
        paste0("95%-CI for beta1: [", round(confint(ols)[2, 1], 2), ",", round(confint(ols)[2, 2], 2), "]"),
        col = "indianred", cex = 1
      )
    }
    abline(
      a = 0, b = true_beta_1,
      lwd = 3, col = "black"
    )
    
    plot(
      y = rep(1:nrow(cis), 3), x = cbind(cis, NA),
      xlim = true_beta_1 + c(-0.6, 0.6), ylim = c(max(50, nrow(cis)), 0), pch = NA,
      xlab = "beta1",
      ylab = "sample",
      main = paste0("CI for beta1. (True beta1 inside CI: ", round(100 * mean(cis[, 1] < true_beta_1 & true_beta_1 < cis[, 2]), 2), "%)"
      )
    )
    lines(
      y = t(cbind(1:nrow(cis), 1:nrow(cis), 1:nrow(cis) )),
      x = t(cbind(cis, NA)),
      lw = 4 / log(log(nrow(cis) + 1) + 1), col = alpha("darkslategray", 0.5)
    )
    abline(v = true_beta_1, col = "black", lw = 4, lty = 'dotted')
    
  }
}
```
